### 2014-11-25 doc/thesis/system.tex
Improving the raw text about the refinement of table types

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
ll = el_{1} \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el_{0} \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el\\
& | \; e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
x \; | \;
{...}_{e} \; | \;
n \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; cl \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
x ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
l ::= & \;\; n \; | \;
e_{1}[e_{2}] \; | \;
n[e] \; {<}t{>}\\
ll ::= & \;\; \vec{l}\\
el_{0} ::= & \;\; \mathbf{nothing} \; | \;
el_{1}\\
el_{1} ::= & \;\; \vec{e} \; | \;
me \; | \;
\vec{e}, me\\
el ::= & \;\; el_{0}\\
cl ::= & \;\; \vec{c} \; | \;
{...}_{m} \; | \;
\vec{c}, {...}_{m}\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
c ::= & \;\; [e_{1}] = e_{2}\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

The most unusual feature of our type system is the refinement of
table types:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
       \env_{1} \vdash x:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
      {\env_{1} \vdash n[x] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
\end{array}
\]

The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
or \emph{unique} table types.
We restricted the keys to literals only because the purpose of
this rule is to make it easier to add new fields to table types
that represent records.
We also do not allow changing the type of a field that is already
in the table because this could lead to loosen table types that
have no static type checking. 
For instance, Typed Lua allows us to write the following code:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$t[``y"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
but it does not allow us to write the following code:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$t[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$t[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}

This example also uses the rule \textsc{T-CONSTRUCTOR1} to
infer the type of the empty table constructor:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Typed Lua also has inference rules to infer table types that we
can use to initialize hashes, arrays, and records:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
       n = |\;\vec{c}\;| \;\;\;
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
      {\env \vdash \{\;\vec{c}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env \vdash {...}:t, \env}
      {\env \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env \vdash c:k_{i}{:}v_{i}, \env_{i} \;\;\;
       \env \vdash {...}:t, \env \;\;\;
       n = |\;\vec{c}\;| \\
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t \cup \Nil\}_{unique})
       \end{array}}
      {\env \vdash \{\;\vec{c},\;{...}_{m}\;\}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}v \cup \Nil\}_{unique}, \env}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR2} type checks the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; x = a[``z"]$}
\end{tabular}
\end{center}

It also uses the rules \textsc{T-FIELD1} and \textsc{T-FIELD2}:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean \cup \Number \cup \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}:super(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       \Nil \not\subtype t_{1}}
      {\env_{1} \vdash [e_{1}] = e_{2}: \Any:super(t_{2}), \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-FIELD1} infers fields that keys are subype
of $\Boolean$, $\Number$, or $\String$.
The rule \textsc{T-FIELD2} infers fields thaty keys have the
dynamic type if its type does not include $\Nil$.

Typed Lua handles arrays as hashes that maps $\Integer$ to some
type $t$.
We did not include in the abstract syntax tree the simplified
syntax of table consturctors that initializes arrays because
we can write it using explicit indices.
For instance, we write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
$\{ ``x", ``y", ``z" \}$.
We chose to make this simplification because the inference rules
for the simpflified version does not include anything new.
They are the same rules that just automatically control the indices.
For the same reason, we also omitted in the table constructor expressions
that return multiple results.
However, we did not omit the vararg expression in the last position
to show the case that we infer a record with an array part.

The rule \textsc{T-CONSTRUCTOR3} type checks the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\Integer:\String \cup \Nil\}_{closed} = \{ {...}_{m} \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\String \cup \Nil = t[1] \; \mathbf{in} \; b = t[5]$}
\end{tabular}
\end{center}

We are assuming that $...$ is in the environment as has type $\String$.

The rule \textsc{T-CONSTRUCTOR4} type checks the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\Integer, ``y":\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m} \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
\end{tabular}
\end{center}

We are also assuming that $...$ is in the environment as has type $\String$.

These rules along \emph{unique} table types also allow us initialize
record types that have optional fields.
We can include the optional field in the initialization
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = ``bar", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}
and we can also ommit it
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Typed Lua also keeps track whether it is safe or not safe to
change the type of a table.
To do that, our type system always promotes an \emph{open}
table to \emph{closed} and an \emph{unique} table to \emph{open}
when they appear as expressions.
The order also matters, as going from \emph{unique} to \emph{open}
before going from \emph{open} to \emph{closed} closes \emph{unique} tables.
The rule \textsc{T-IDREAD} along with the predicate \emph{close}
express this intuition:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env_{1}(n_{r}) = t}
      {\env_{1} \vdash n_{r}:close(t), \env_{1}}
\end{array}
\]

The rule \textsc{T-IDREAD} makes this unsafe example not type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``x":\String \cup \Nil\}_{closed} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil}$}
\end{tabular}
\end{center}

@@

However, sometimes we want to use \emph{unique} tables to initialize
closed tables, as they allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use the \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{t \subtype \env_{1}(n)}
      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
\end{array}
\]

@@

\textit{/* Expression list rules. Where should we describe the
discarding/replacement of tuple elements? */}

>>>>

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
c \; | \;
{...}_{e} \; | \;
n \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; {...}_{m} \; \} \; | \;
\{ \; \vec{p},{...}_{m} \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
l ::= & \;\; n \; | \;
e_{1}[e_{2}] \; | \;
n[c] \; {<}t{>}\\
c ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

The most unusual feature of our type system is the refinement of
table types.
The rule \textsc{T-REFINE} allow us to add new keys to \emph{open}
and \emph{unique} table types:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\env_{1}(n) = \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{open|unique} \;\;\;
       \env_{1} \vdash c:t_{1}, \env_{2} \;\;\;
       \not \exists i \in 1..n \; t_{1} \lesssim k_{i}}
      {\env_{1} \vdash n[c] {<}t{>}:t, \env_{2}[n \mapsto \{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, t_{1}{:}t\}_{open|unique}]}
\end{array}
\]

We restricted the refinement of table types to include only literal
keys because its purpose is to make it easier the construction of
table types that represent records.
For instance, 
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$a[``y"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
uses field assignment to add fields $x$ and $y$ to the local $a$.
After these field assignments $a$ has type
$\{``x":\String, ``y":\Integer\}_{unique}$.

We do not allow the refinement of table types to change the type of
a field that is already present in the table type.
For instance,
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
do not type check, as we are trying to change the type of field
``x" from $\String$ to $\String \cup \Integer$.
More precisely, our type system does not allow the table type to
change towards a type that is not a subtype of the previous type.
In this example,
$\{``x":\String \cup \Integer\}_{unique} \not\subtype \{``x":\String\}_{unique}$.
We have this restriction because this kind of refinement could lead
to fields that have a too general type.

\textit{/* We need to close table types before assigning them to fields. */}
\begin{center}
\begin{tabular}{l}
$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in} \; a[``x"] \; {<}\{\}_{closed}{>} = \{\}$
\end{tabular}
\end{center}

Lua has multiple assignments, so Typed Lua also has to take this into
consideration while refining table types.
The rule \textsc{T-ASSINGMENT} shows that tuple types make it easy
to use the consistent-subtyping relation to check multiple assignments:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \vec{l} = el:\env_{3}}
\end{array}
\end{array}
\]

This rule is so simple because we left the difficult part to the
type checking of expression lists.
The list of left-hand values is also a kind of expression list.
The difficult part is to check whether there is an expression in
the expression list that messes the environment, and the rule
\textsc{T-LHSLIST} captures this intuition:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env \vdash l_{k}:t_{k}, \env_{k} \;\;\;
       merge(\env_{1}, ..., \env_{n}) = \env_{m} \;\;\;
       n = |\;\vec{l}\;|}
      {\env \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
to the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.

We can also use multiple assignments to refine table types:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``y"] \; {<}\Integer{>} = ``foo", 1$}\\
\end{tabular}
\end{center}

This example type checks because all the environment changes are consistent, and
$``foo" \times 1 \times \Nil{*} \lesssim \String \times \Integer \times \Value{*}$.
The type of the expression list includes the type $\Nil{*}$ in the end
because our type system has typing rules to add this type when
the type of the expression list does not end with a variadic type.
Our type system has this behavior to use $\Nil$ in the place of missing
values, like Lua does.
This example uses the rule \textsc{T-EXPLIST2}, which is really close
to the rule \textsc{T-LHSLIST}:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
       merge(\env_{1}, ..., \env_{n}) = \env_{m} \;\;\;
       n = |\;\vec{e}\;|}
      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

The next example does not type check because it tries to add
the same field to $a$:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``x"] \; {<}\Integer{>} = ``foo", 1$}\\
\end{tabular}
\end{center}

Typed Lua also keeps track whether it is safe or not safe to
change the type of a table, but before discussing this subject,
we will discuss the different table constructors that our
abstract syntax includes, as they will appear in the next
examples.

All the examples that we presented until now use the simplest form of
the table constructor: the empty table $\{\}$.
Its inference rule is starightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of expression pairs $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions because we can write it using the first form.
For instance, we can write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
$\{ ``x", ``y", ``z" \}$.
We chose to not include this form because its inference rule is similar
to the inference rule of the first form, and just controls the type of
the ommited indices.
For the same reason we did not include table constructors with
multiple expressions that can return a tuple type that does not end
with a variadic type.
We did include the vararg expression in the last position to show how
our type system can infer a table type that has a record part and
also an array part.

The inference rule \textsc{T-CONSTRUCTOR2} infers a table type for
the first form:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}) = t}
      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
\end{array}
\]

This inference rule infers the type of each pair, uses these types
to build a table type, and uses the predicate \emph{wf} to check
whether the table type is well formed.
We gave the definition of this predicate in the last section.

The inference rules for expression pairs would be straightforward
if our table types allowed any type in the keys.
The rules \textsc{T-PAIR1}, \textsc{T-PAIR2}, and \textsc{T-PAIR3}
check if the type of the key is a $\Boolean$, a $\Number$,
or a $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-PAIR1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR3}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-PAIR4} is a fallback for the previous rules,
but it works only if the type of the key does not include $\Nil$:
\[
\begin{array}{c}
\mylabel{T-PAIR4}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       \Nil \not\subtype t_{1}}
      {\env_{1} \vdash [e_{1}] = e_{2}: \Any{:}t_{2}, \env_{3}}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR2} along with the rules for expression
pairs allow Typed Lua to type check the following example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``x"] = 1, [``y"] = 2 \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\Integer \cup \Nil = a[``x"] \; \mathbf{in} \; b = a[``z"]$}
\end{tabular}
\end{center}

This example type checks because Typed Lua infers the table type
$\{``x":1, ``y":2\}_{unique}$, which is subtype of
$\{\String:\Integer \cup \Nil\}_{closed}$.
Even though the key $``z"$ does not exist in the table $a$,
the two table indexations type check with type $\Integer \cup \Nil$,
which is subtype of the type of $b$.
The rule \textsc{T-INDEX1} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
\end{array}
\]

Typed Lua handles arrays as hashes that maps integers to some type $t$.
In Lua, programmers often use the vararg expression to initialize arrays.
The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
the behavior of the vararg expression for this case:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env_{1}({...}) = t}
      {\env_{1} \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env_{1}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       \env_{i+1}({...}) = t_{v}\\
       wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique}) = t_{t}
       \end{array}}
      {\env_{1} \vdash \{\;\vec{p},\;{...}_{m}\;\}:t_{t}, \env_{n+1}}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
only the vararg expression inside a table constructor to initialize
an array.
If we assume that the $...$ is in the environment and has type $\String$,
the following example type checks through this rule:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\Integer:\String \cup \Nil\}_{closed} = \{ {...}_{m} \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; b:\String \cup \Nil = t[1] \; \mathbf{in} \; b = t[5]$}
\end{tabular}
\end{center}

The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
the table constructor as a record that includes an array part.
If we assume that the $...$ is in the environment and has type $\String$,
the following example type checks through this rule:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\Integer, ``y":\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = \mathbf{true}, {...}_{m} \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
\end{tabular}
\end{center}
but the next example does not type check because the record part
includes fields that the key types are subtype of the array part:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{1:\String, 10:\Boolean, \Integer:\String \cup \Nil\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [1] = ``foo", [10] = \mathbf{true}, {...}_{m} \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; a[``x"] = ``bar"$}
\end{tabular}
\end{center}

Making the inference rules of the table constructor infer an \emph{unique}
table type also allow us to use them for initializing record types that
have optional fields.
We can include the optional field in the initialization
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``y"] = ``bar", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}
and we can also omit it in the initialization
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{``x":\String, ``y":\String \cup \Nil, ``z":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``x"] = ``foo", [``z"] = ``baz" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

After discussing the inference rules of the table constructor,
we can get back to the discussion about the refinement of table types.

Besides allowing the refinement of table types through field assignment,
Typed Lua also keeps track whether it is safe or not safe to change the
type of a table.
To do that, our type system always promotes an \emph{open}
table to \emph{closed} and an \emph{unique} table to \emph{open}
when they appear as expressions, or when they enter in a scope that is
not its declaration scope.
The order matters, as going from \emph{unique} to \emph{open}
before going from \emph{open} to \emph{closed} closes \emph{unique} table types.

The rule \textsc{T-IDREAD} uses the predicate \emph{close} to do
this promotion before we use a local variable:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env_{1}(n_{r}) = t}
      {\env_{1} \vdash n_{r}:close(t), \env_{1}}
\end{array}
\]

This rule does not change the type of the local variable because it
just need to ensure that an alias will not get the same type.
For instance, it makes this unsafe example not type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``x":\String \cup \Nil\}_{closed} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil}$}
\end{tabular}
\end{center}

@@

However, sometimes we want to use \emph{unique} table types to initialize
\emph{closed} table types, as they allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{t \subtype \env_{1}(n)}
      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
\end{array}
\]

@@
